Enhance Road Detection Data Processing of LiDAR Point Clouds to Specifically Identify Unmarked Gravel Rural Roads

### 2023-04-27 Huston_Paper.tex
okay here we go

	\author{Rhett Huston, Jay Wilhelm}
%	{
%		and~Jane~Doe,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
%		\thanks{M. Shell was with the Department
%			of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
%			GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
%		\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
%		\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}}

@@

% The paper headers
\markboth{Some Sort of Journal}%
{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for IEEE Journals}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}

	{Gravel and chipseal roads that lack standardized features such as curbs or painted lines present detection and navigation challenges to autonomous vehicles. Differentiation between road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation. Global Positioning Service (GPS) and high resolution maps cannot be relied upon for navigation of rural roads, as some may only be width of the vehicle and GPS may not be accurate enough. Normal Distribution Transform (NDT) based on scanning LiDAR data may be insufficient for navigating on gravel roads. This work will examine a method of classifying scanning LiDAR data spatial and remission features for explicit detection of unmarked gravel and chipseal road surfaces. Exploration of terrain classification using high resolution scanning LiDAR data of these road surfaces will allow for predicting road boundary location potentially enabling confident autonomous operations on gravel roads. The principal outcome of this work is a method for gravel road detection using LiDAR data for the purpose of predicting road boundary locations. Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.}

\end{abstract}

@@

	% Autonomous vehicles have difficulty detecting rural unpaved gravel roads due to lack of standardized features such as curbs and painted lane lines. 
	%	Autonomous vehicles have relied upon standardized road features such as curbs and painted lane lines or high definition maps paired with precision GPS for navigational solutions.

@@

	{LiDAR has been used by autonomous vehicles for road edge detection purposes, however challenges arise in environments with unpaved gravel and chipseal roads due to the lack of standardized road features such as curbs and painted lane lines. Gravel roads comprise 34.8\% \cite{road_stats_2} of all road surfaces in the United States. Some nations have predominately unpaved road networks, such as India, where 70.7\% of all roads by mile are categorized as unpaved \cite{malik_lal_2019}. Detection of unmarked roads may allow for expansion of current civilian autonomous vehicle capabilities as well as introduce opportunities for industries that operate in rural areas. Global Positioning Service (GPS) and high resolution maps cannot be relied upon for navigation of rural roads, as some gravel roads may only be width of the vehicle and GPS is not accurate enough or may not have a reliable signal for precision driving \cite{noauthor_gpsgov_nodate}. Systems that rely on cameras may fail when analyzing gravel roads, as these lack visual cues such as painted lane markings \cite{crisman_scarf_1993} and may in some circumstances closely resemble surrounding terrain. LiDAR point cloud processing systems depend upon distinct geometric features, most commonly curbs \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_nodate,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja_road_nodate}, however as curbs are not installed on rural gravel roads \cite{skorseth_gravel_nodate} this method breaks down. Normal Distribution Transform (NDT) scan matching analytically compares two point cloud data sets to track current position, however this model relies on distinct geometric features, and suffers when distinguishing between terrain types making it insufficient for determining road boundaries \cite{biber_normal_2003}. Using 2D projection of planes unto a 3D point cloud of a road surface is an alternative approach, but likewise requires curbs or painted lines for road boundary definition \cite{fernandes_road_2014, borkar_robust_2009-1, guo_lane_2015}. Instead, using a LiDAR forward approach may overcome difficulties in achieving real-time detection of unmarked roads by using terrain classification to discover road type, boundaries, and intersections.}

@@

	{Detection of unmarked road surfaces may rely on the roughness properties without relying on distinct markings or topographical features. Surface roughness is a measurable property that may be used to characterize a gravel road, as they are typically consistent with gravel type used \cite{skorseth_gravel_nodate} which may be exploited in a search and compare function, and is distinct from grass and other ground types \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain}. Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}. Implementing a terrain classification model using Random Decision Forests (RDF) using scanning LiDAR data to exploit surface roughness was studied for road surface detection in this study.}

	{Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident. Problem Statement: Determine a processing method for increasing road detection precision of unmarked gravel rural roads.}

@@

	{MLS and RANSAC were compared as methods to project a two dimensional plane unto a point cloud, facilitating examination of points by providing a frame of reference \cite{miller_method_nodate, gojcic_perfect_2019}. Method of least-squares analyses uses regression to approximate the best fit to a data set (\ref{eqn:MLS}). RANSAC is an iterative algorithm that is designed to handle large numbers of outliers in the sample data set \cite{derpanis_overview_nodate,yaniv_random_2010,fischler_random_1987,cantzler_random_nodate}. 
	RDFs are a supervised machine learning method that operates on a set of decision trees \cite{ho_random_1995}. REFs have been shown to be useful with handling LiDAR data sets \cite{breiman_random_2001}, and have been employed in terrain classification \cite{laible_3d_2012,laible_map_building,laible_terrain_2013,khan_high_2011,reymann_improving_2015,schilling_geometric_2017, wietrzykowski_context-aware_2019}. Multiple decision trees are created with bootstrapped data sets - randomly selected data sub-sets from a training set. Classification was determined by the majority results as opposed to the mean \cite{breiman_random_2001,ho_random_1995}. RDFs may be used to predict road surface area by detecting gravel terrain.}

@@

	{Required physical data was gathered by the Hexegon Chrysler Pacifica New Eagle DBW van (Figure \ref{fig:van}), built by AutonomouStuff, owned and operated by Ohio University. LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C (Figure \ref{fig:vlp32mount}). Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a high output data stream \cite{vlp_32c}. GPS data was recorded with a PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS) manufactured by Novatel, and two GNSS-502 antennas by NavtechGPS. Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using Robotic Operating System (ROS). Scanning LiDAR training data was collected on various roads and lots near Athens, Ohio. Scanning LiDAR data of gravel surfaces was gathered principally from an un-named gravel road near Armig Road (Figure \ref{fig:Armig_Road_Camera_View})and a gravel lot maintained by the Ohio University Civil Engineering Department near Blackburn Road. Additional gravel road data for testing purposes only was gathered on Sturbois Road. Scanning LiDAR data of chipseal surfaces was gathered principally from Bean Hollow Road. Additional chipseal road data for testing purposes only was gathered on later portions of both Bean Hollow Road (Figure \ref{fig:Bean_Cam_View}). Scanning LiDAR data of grassy surfaces and foliage was gathered from the sides of the previously stated roads, therefore no special data collection drive was made for those terrain types.}

	\begin{figure}[H] 
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/Van}
			\caption[Sensor Van]{}
			\label{fig:van}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/vlp_32_mount_2}
			\caption[VLP 32 on Van]{}
			\label{fig:vlp32mount}
		\end{subfigure}
		\caption[Experimental Apparatus]{Experimental Apparatus (a). Velodyne VLP-32 Scanning LiDAR is mounted on top of the vehicle (b).}
		\label{fig:Experimental_Apperatus}
	\end{figure}

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/Bean_Hollow_Road_Camera}
			\caption[Bean Hollow Road Camera View]{}
			\label{fig:Bean_Cam_View}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/armig_2}
			\caption[Armig Road Camera View]{}
			\label{fig:Armig_Road_Camera_View}
		\end{subfigure}
		\caption[Armig Road \& Bean Hollow Road]{Typical examples of unmarked gravel and chipseal roads. Road view of Bean Hollow Road - an unmarked chipseal road (a). Road view of Armig Road - an unmarked gravel road (b).}
		\label{fig:Combined_Roads}
	\end{figure}

	{MATLAB was used to process the gathered rosbag files using the MATLAB ROSBAG library. Rosbags were directly loaded into the MATLAB environment. Point cloud topic was defined, and the point cloud messages were read from the topic and imported as a data structure. Cartesian coordinates and corresponding intensity values were extracted and saved to disk. GPS and IMU data was extracted in a similar manner. GPS and IMU topics were defined, and topic messages are read. Longitude, latitude, and altitude were extracted from the GPS messages, while roll, pitch, and yaw were extracted from the IMU messages.}

	{Scanning LiDAR training and verification data was extracted by selecting appropriate consecutive points from a single LiDAR channel lying on gravel, chipseal, foliage, and grass surfaces. Number of points included in each training data set was not controlled. Foliage was defined as trees and bushes, while grass was defined as a trimmed lawn. Projection of a plane was necessary to provide a reference point for describing point variance. RANSAC and Method of Least Squares were tested for use as plane projection methods. Additionally, a point's perpendicular distance as well as range from the LiDAR origin (Figure \ref{fig:xy_vs_range}) was also considered, both of which has a computational advantage over the plane projection method in that they may be read directly from the Velodyne VLP-32 data stream.}

	{RANSAC was implementing by a built-in MATLAB function, $pcfitplane$. Method of Least Squares was implemented by importing a MATLAB library \cite{noauthor_object-oriented_nodate}. Point clouds are individually imported into the MATLAB environment. RANSAC or Method of Least Squares may be used to project a reference plane unto the point cloud. Displaying the point cloud is completed using MATLAB's $pcshow$ function. Training data is selected by manually drawing an area around appropriate points of a desired terrain type using MATLAB's $drawrectangle$ function (Figure \ref{fig:Rectangle_ROI}). Indices for the rectangle are extracted, and MATLAB's function $findPointsInROI$ is used to extract sympatric points within the drawn rectangle (Figure \ref{fig:ROI_Points}). Captured points are then saved to disk along with the projected plane's normal vector and distance from the origin $[a,b,c,d]$.}

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.5\linewidth]{figures/Rectangle_ROI}
			\caption[Region of Interest Selection]{MATLAB was used to select training data using the $drawrectangle$ function and extracting points that lay within the defined area.}
			\label{fig:Rectangle_ROI}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.5\linewidth]{figures/ROI_Points}
			\caption[Points in Region of Interest]{Selected points (Figure \ref{fig:Rectangle_ROI}) lying on an unmarked road.}
			\label{fig:ROI_Points}
		\end{subfigure}
		\caption[Armig Road \& Bean Hollow Road]{Road view of Bean Hollow Road - an unmarked chipseal road (a). Road view of Armig Road - an unmarked gravel road (b).}
		\label{fig:Combined_Things}
	\end{figure}

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.5\linewidth]{figures/xy_vs_range}
		\caption[XY vs Range vs Z Height]{Three references for feature extraction may be seen in this simplified graph. First is the Range shown by the magenta line from the LiDAR origin to the Point of Interest (PoI). Second is the height from  LiDAR origin to the PoI (cyan). Finally, distance from the projected plane (red) to the PoI (green). XY Distance (blue) and Range (magenta) from the sensor was used consider the effect of distance from LiDAR origin to the PoI.}
		\label{fig:xy_vs_range}
	\end{figure}

	{Feature extraction was completed by performing mathematical functions on the gathered spatial and remission training data. Reference plane normal vector and distance from origin $[a,b,c,d]$ data derived from either RANSAC or Method of Least Squares methods was selected. Base-line features extracted include the following: Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial or remission data array. Distance from the LiDAR point of origin was also considered by dividing the base-line features by either the average range or X-Y distances for a set of points, as LiDAR accuracy may decrease as distance increases. Sixty features were considered in total for this work, and multiple feature sub-sets were used to create a variety of Random Decision Forests. Plane-projection based algorithms were created considering all features. Determination of feature usage allowed for creation of RDFs that used only the top twenty features. Non-plane projection based algorithms were created considering only range and Z height features. For all RDFs, both spatial and remission features were used. }

	{Random Decision Forests were created using MATLAB's $TreeBagger$ function, which creates an ensemble of bagged decision trees using the input arguments of included features, tree depth, number of trees, and a cost array. RDFs were split into sub categories within each feature set to test varying tree depths (Figure \ref{fig:all_tree_comp}), and allowed to grow in number of trees until three hundred trees were created. Discouragement of false-positive identification for road surfaces (gravel and chipseal) was accomplished by introducing a cost array which would induce a bias against gravel and chipseal . Validation error was used as the primary means to compare RDF performance prior to testing with data outside the training and validation data sets. MATLAB was used to load verification data and test increasing sizes of RDF algorithms for accuracy by comparing the known data types against the RDF result. Plotting number of trees verses error allowed for determination of an early stopping point for algorithm growth (Figure \ref{fig:mls_zxy_early_stopping}) by choosing the point where the change in error moving average did not exceed 10\% and was within 20\% of the minimum error.}

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{figures/MLS_RAN_ALL}
			\caption[]{}
			\label{fig:all_mls_ransac_comp}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{figures/MLS_RAN_TT}
			\caption[]{}
			\label{fig:xyz_mls_ransac_comp}
		\end{subfigure}
		\label{fig:xyz_mls_comp_1}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{figures/Range_Vali_Error}
			\caption[]{}
			\label{fig:range_mls_ransac_comp}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{figures/ZXY}
			\caption[]{}
			\label{fig:t20_mls_ransac_comp}
		\end{subfigure}
		\caption[RDF Validation Accuracy]{RDF validation accuracy directly compared. All features are considered (a); Top 20 features used in (a) only (b); Range was used without X, Y, and Z components broken out (c); Plane projection not considered in (d), reference plane set as the LiDAR origin point.}
		\label{fig:all_tree_comp}
	\end{figure}

	\begin{figure}[H]
			\centering
			\includegraphics[width=0.75\linewidth]{figures/mls_zxy_early_stopping}
			\caption[Early Stopping]{Diminishing returns on increasing algorithm size versus accuracy can be used to determine the early stopping point of algorithm growth.}
			\label{fig:mls_zxy_early_stopping}
	\end{figure}

@@

\section{All Results}\label{apdx:Appendix_2}

\begin{table}[]
	\begin{tabular}{|>{\centering\arraybackslash}p{.25cm}|>{\centering\arraybackslash}p{.25cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.4cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|}
		\hline

>>>>

	\author{Rhett Huston, Dr. Jay Wilhelm}
%	{
%		and~Jane~Doe,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
%		\thanks{M. Shell was with the Department
%			of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
%			GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
%		\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
%		\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}}

@@

% The paper headers
\markboth{Journal of Autonomous Vehicles and Systems}%
{}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}{

	Gravel roads lack standardized features such as curbs or painted lines, presenting detection challenges to autonomous vehicles. Differentiation between road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation. Global Positioning Service (GPS) and high resolution maps may not be reliable for navigation of gravel roads, as some may only be width of the vehicle and GPS may not be accurate enough. Normal Distribution Transform (NDT) based on scanning LiDAR data may be insufficient for navigating on gravel roads as there may not be enough geometrically distinct features for reliable scan matching. This work will examine a method of classifying scanning LiDAR data spatial and remission features for explicit detection of unmarked gravel road surfaces. Exploration of terrain classification using high resolution scanning LiDAR data of these road surfaces may allow for predicting gravel road boundary locations potentially enabling confident autonomous operations on gravel roads. The principal outcome of this work was a method for gravel road detection using LiDAR data for the purpose of predicting road boundary locations. Random Decision Forests were tested for the purpose of scanning LiDAR data terrain classification. Manually discovered gravel and paved road surface based on classified results were compared to actual road bounds. <SUMMARY OF RESULTS HERE>. Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.
}	
\end{abstract}

@@

	% Autonomous vehicles have difficulty detecting rural unpaved gravel roads due to lack of standardized features such as curbs and painted lane lines. 
	%	Autonomous vehicles have relied upon standardized road features such as curbs and painted lane lines or high definition maps paired with precision GPS for navigational solutions. 
	% LiDAR has been used by autonomous vehicles for road edge detection purposes, however challenges arise when trajectory solutions are required on unpaved gravel and chipseal roads due to the lack of standardized road features such as curbs and painted lane lines.

@@

	{Gravel roads comprise 34.8\% \cite{road_stats_2} of all road surfaces in the United States. Some nations have predominately unpaved road networks, such as India, where 70.7\% of all roads by mile are categorized as unpaved \cite{malik_lal_2019}. Detection of gravel roads may allow for expansion of current civilian autonomous vehicle capabilities as well as introduce opportunities for industries that operate in rural areas. Global Positioning Service (GPS) and high resolution maps cannot be relied upon for navigation of rural roads alone as GPS is not accurate enough or may not have a reliable signal for precision driving \cite{noauthor_gpsgov_nodate}, and gravel roads are not marked or mapped. Systems that rely on cameras may fail when analyzing gravel roads, as these lack visual cues such as painted lane markings \cite{crisman_scarf_1993} and may in some circumstances closely resemble surrounding terrain. LiDAR point cloud processing systems depend upon distinct geometric features, most commonly curbs \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_nodate,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja_road_nodate}, however as curbs are not installed on rural gravel roads \cite{skorseth_gravel_nodate} this method breaks down. Normal Distribution Transform (NDT) scan matching analytically compares two point cloud data sets to track current position, however this model relies on distinct geometric features, and suffers when distinguishing between terrain types making it insufficient for determining road boundaries \cite{biber_normal_2003}. Using 2D projection of planes unto a 3D point cloud of a road surface is an alternative approach, but likewise requires curbs or painted lines for road boundary definition \cite{fernandes_road_2014, borkar_robust_2009-1, guo_lane_2015}. Instead, using LiDAR may overcome difficulties in achieving real-time detection of unmarked roads by using terrain classification to discover road type, boundaries, and intersections.}

@@

	{Detection of unmarked road surfaces may utilize roughness properties without distinct markings or topographical features. Surface roughness is a measurable property that may be used to characterize a gravel road, as they are typically consistent with gravel type used \cite{skorseth_gravel_nodate} which may be exploited in a search and compare function, and is distinct from chipseal, grass, and other ground types \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain}. Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}. Implementing a terrain classification model using Random Decision Forests (RDF) using scanning LiDAR data to exploit surface roughness was studied for road surface detection in this study. Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident. This work will attempt to determine a processing method for increasing road detection precision of unmarked gravel rural roads.}

@@

%	{MLS and RANSAC were compared as methods to project a two dimensional plane unto a point cloud, facilitating examination of points by providing a frame of reference \cite{miller_method_nodate, gojcic_perfect_2019}. Method of least-squares analyses uses regression to approximate the best fit to a data set (\ref{eqn:MLS}). RANSAC is an iterative algorithm that is designed to handle large numbers of outliers in the sample data set \cite{derpanis_overview_nodate,yaniv_random_2010,fischler_random_1987,cantzler_random_nodate}.}

	{Random Decision Forests are a supervised machine learning method that operates on a set of decision trees \cite{ho_random_1995}. RDFs have been shown to be useful with handling LiDAR data sets \cite{breiman_random_2001}, and have been employed in terrain classification \cite{laible_3d_2012,laible_map_building,laible_terrain_2013,khan_high_2011,reymann_improving_2015,schilling_geometric_2017, wietrzykowski_context-aware_2019}. Multiple decision trees are created with bootstrapped data sets - randomly selected data sub-sets from a training set. Classification is determined by the majority results as opposed to the mean \cite{breiman_random_2001,ho_random_1995}. Bayesian Optimization may be used to tune the hyper parameters of a machine algorithm \cite{noauthor_bayesian_nodate, snoek_practical_2012}.}

	% Summary of lit review in 1 sentance:
	{Current methods for road detection may be insufficient for real-time detection of unmarked gravel and chipseal roads due to ambient conditions, lack of standardized features, or rely on post-processing large data sets, while terrain classification of scanning LiDAR data may overcome these difficulties.}

@@

	{Required physical data was gathered by the Ohio University Autonomous Van van (Figure \ref{fig:van}), built by AutonomouStuff. LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C (Figure \ref{fig:vlp32mount}). Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a high output data stream \cite{vlp_32c}. GPS data was recorded with a Novatel PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS), and two GNSS-502 antennas by NavtechGPS. Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using Robotic Operating System (ROS). Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot that intersects Blackburn Road near Athens Ohio. Scanning LiDAR data of an asphalt surface was gathered from Blackburn Road near the intersecting gravel lot. Scanning LiDAR data of grassy surfaces was gathered from the side of the gravel lot. Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection. Only three classes are considered in this work: "gravel", "asphalt", and "unknown". }

	\begin{figure}[H] 
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/van_on_van}
			\caption[Sensor Van]{}
			\label{fig:van}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/LiDAR_on_van}
			\caption[VLP 32 on Van]{}
			\label{fig:vlp32mount}
		\end{subfigure}
		\caption[Experimental Apparatus]{Experimental Apparatus (a). Velodyne VLP-32 Scanning LiDAR is mounted on top of the vehicle (b).}
		\label{fig:Experimental_Apperatus}
	\end{figure}

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/vlcsnap-2023-04-20-08h48m17s447}
			\caption[Blackburn Road Camera View]{}
			\label{fig:Bean_Cam_View}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/armig_2}
			\caption[Armig Road Camera View]{}
			\label{fig:Armig_Road_Camera_View}
		\end{subfigure}
		\caption[Blackburn Road \& Gravel Parking Lot]{Road view of Blackburn Road - an unmarked asphalt road (a). Road view of the gravel driveway (b).}
		\label{fig:Combined_Roads}
	\end{figure}

	{MATLAB was used to process the gathered rosbag files using the their ROSBAG library. Rosbags were directly loaded into the MATLAB environment. Point cloud, GPS, and IMU topics were defined, and the messages were read from the topic and imported. Cartesian coordinates, longitude, latitude, altitude, roll, pitch, and yaw were extracted from the messages.}

	{Training data was extracted by manually defining gravel, asphalt, and unknown terrain areas on compiled point clouds. Compiling LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road. Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data. Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU. LiDAR origin was found using GPS longitude, latitude, and altitude data. Physical distances between the GPS, IMU, and LiDAR was rectitude by obtaining reference frames provided by AutonomouStuff. GPS coordinates were offset by the current orientation and converts the ground truth to the LiDAR frame. GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors. Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}). While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for scoring purposes. Manually defined areas may then be overlayed unto the point cloud (Figure \ref{fig:Combined_Classified_PCD}). Individual scans were consecutively examined to extract per-channel scanning LiDAR training data and exported to a database if coincident with the manually defined areas.}

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.95\linewidth]{figures/Compiled_PCD}
			\caption[C]{}
			\label{fig:Compiled_PCD}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.95\linewidth]{figures/Manually_Defined_Area_Train_Dat_Grab_2}
			\caption[]{}
			\label{fig:Classified_Areas_PCD}
		\end{subfigure}
		\caption[Compiled and Classified Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map (a). Manually defined areas inform where training data lies (b).}
		\label{fig:Combined_Classified_PCD}
	\end{figure}

%	{RANSAC was implementing by a built-in MATLAB function, $pcfitplane$. Method of Least Squares was implemented by importing a MATLAB library \cite{noauthor_object-oriented_nodate}. Point clouds are individually imported into the MATLAB environment. RANSAC or Method of Least Squares may be used to project a reference plane unto the point cloud. Displaying the point cloud is completed using MATLAB's $pcshow$ function. Training data is selected by manually drawing an area around appropriate points of a desired terrain type using MATLAB's $drawrectangle$ function (Figure \ref{fig:Rectangle_ROI}). Indices for the rectangle are extracted, and MATLAB's function $findPointsInROI$ is used to extract sympatric points within the drawn rectangle. Captured points are then saved to disk along with the projected plane's normal vector and distance from the origin $[a,b,c,d]$.}

%	\begin{figure}[H]
%		\centering
%		\begin{minipage}[c]{.45\linewidth}
%			\includegraphics[width=.95\linewidth]{figures/Rectangle_ROI}
%			\caption[Region of Interest Selection]{MATLAB was used to select training data using the $drawrectangle$ function and extracting points that lay within the defined area.}
%			\label{fig:Rectangle_ROI}
%		\end{minipage}
%		\hfill
%		\begin{minipage}[c]{0.45\linewidth}
%			\includegraphics[width=.95\linewidth]{figures/xy_vs_range}
%			\caption[XY vs Range vs Z Height]{Three references for feature extraction may be seen in this simplified graph. First is the Range shown by the magenta line from the LiDAR origin to the Point of Interest (PoI). Second is the height from  LiDAR origin to the PoI (cyan). Finally, distance from the projected plane (red) to the PoI (green). XY Distance (blue) and Range (magenta) from the sensor was used consider the effect of distance from LiDAR origin to the PoI.}
%			\label{fig:xy_vs_range}
%		\end{minipage}
%	\end{figure}

	{Feature extraction was completed by performing mathematical functions on the gathered spatial and remission training data (Figure \ref{fig:flowzchart}). Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial or remission data array. Features were non-dimensionalized as necessary by dividing by the average distance to the appropriate power to cancel units. Thirty percent of the training data was randomly selected and reserved to create a validation data set.}

	{MATLAB's Classification Learner Application was used to generated Random Decision Forests. Decision tree depth and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization.}

	{Random Decision Forests were created using MATLAB's $TreeBagger$ function, which creates an ensemble of bagged decision trees using the input arguments of included features, tree depth, number of trees, and a cost array. RDFs were split into sub categories within each feature set to test varying tree depths (Figure \ref{fig:all_tree_comp}), and allowed to grow in number of trees until three hundred trees were created. Discouragement of false-positive identification for road surfaces (gravel and chipseal) was accomplished by introducing a cost array which would induce a bias against gravel and chipseal. Validation error was used as the primary means to compare RDF performance prior to testing with data outside the training and validation data sets. MATLAB was used to load verification data and test increasing sizes of RDF algorithms for accuracy by comparing the known data types against the RDF result. Plotting number of trees verses error allowed for determination of an early stopping point for algorithm growth (Figure \ref{fig:mls_zxy_early_stopping}) by choosing the point where the change in error moving average did not exceed 10\% and was within 20\% of the minimum error.}

	\begin{figure}
		\centering
		\includegraphics[width=0.9\linewidth]{figures/flowz}
		\caption[Project Flow]{Project flowchart}
		\label{fig:flowzchart}
	\end{figure}

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{figures/RAN_MLS_ALL}
			\caption[]{}
			\label{fig:all_mls_ransac_comp}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{figures/RAN_MLS_TT}
			\caption[]{}
			\label{fig:xyz_mls_ransac_comp}
		\end{subfigure}
		\label{fig:xyz_mls_comp_1}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{figures/Range}
			\caption[]{}
			\label{fig:range_mls_ransac_comp}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{figures/zxy}
			\caption[]{}
			\label{fig:t20_mls_ransac_comp}
		\end{subfigure}
		\caption[RDF Validation Accuracy]{RDF validation accuracy directly compared. All features are considered (a); Top 20 features used in (a) only (b); Range was used without X, Y, and Z components broken out (c); Plane projection not considered in (d), reference plane set as the LiDAR origin point.}
		\label{fig:all_tree_comp}
	\end{figure}

	\begin{figure}[H]
			\centering
			\includegraphics[width=0.75\linewidth]{figures/early_stopping_example}
			\caption[Early Stopping]{Diminishing returns on increasing algorithm size versus accuracy can be used to determine the early stopping point of algorithm growth.}
			\label{fig:mls_zxy_early_stopping}
	\end{figure}

@@
